Evolution Reinforces Cooperation with the Emergence of Self-Recognition Mechanisms: an empirical study of the Moran process for the iterated Prisoner's dilemma

### 2017-08-09 main.tex
Fix the details about the training.

This manuscript provides a detailed numerical analysis of agent-based simulations
of \textbf{\input{./tex/num_strategies.tex}}complex and adaptive strategies for the
IPD\@. This is made possible by the Axelrod library \cite{axelrodproject}, an
effort to provide software for reproducible research for the IPD\@. The library
now contains over \input{./tex/num_strategies_axelrod.tex}parameterized
strategies including classics like TitForTat and WinStayLoseShift, as well as
recent variants such as OmegaTFT, zero determinant and other memory one
strategies, strategies based on finite state machines, lookup tables, neural
networks, and other machine learning based strategies, and a collection of novel
strategies. Not all strategies have been considered for this study: excluded
are those that make use of knowledge of the number of turns in a match
and others that have a high
computational run time. The large number of strategies is available thanks to
the open source nature of the project with over 40 contributors from around the
world, made by programmers and researchers \cite{Knight2016}. Three of the considered
strategies are finite state machines trained specifically for Moran processes
(described further in Section~\ref{sec:strategies}).

@@

\begin{itemize}
	\item TF1 \(N=12\), 0\% noise, 10000 repetitions per matchup
	\item TF2 \(N=10\), 0\% noise, 10000 repetitions per matchup
	\item TF3 \(N=8\), 1\% noise, 100 repetitions per matchup
\end{itemize}

Each matchup of players was run to fixation for the specified number of
repetitions to estimate the absorption probabilities. The trained algorithms
were run for fewer than 50 generations. Training data for this is available at
\cite{data}.

@@

Each strategy pair is run for 1000 repetitions of the Moran process to fixation
with starting population distributions of $(1, N-1)$, $(N/2, N/2)$ and $(N-1 ,
1)$, for \(N\) from 2 through 14. The fixation probability is then empirically
computed for each combination of starting distribution and value of \(N\).  The
Axelrod library can carry out exact simulations of the Moran process. Since some
of the strategies have a high computational cost or are stochastic, samples are
taken from a large number of match outcomes for the pairs of players for use in
computing fitnesses in the Moran process. This approach was verified to agree
with unsampled calculations to a high degree of accuracy in specific cases.
This is described in Algorithms~\ref{alg:data_collection}
and~\ref{alg:moran_process}.

@@

It can be seen that apart from CS, none of the strategies
of Table~\ref{tbl:summary_top_2} performs well for \(N\in\{3, 7, 14\}\). The new
top performing strategies are:

@@

Table~\ref{tbl:top_resist} shows the top strategies when ranked
according to \(x_{N-1}\) for \(N\in\{3, 7, 14\}\).
Once again none of the short memory strategies from
Section~\ref{sec:two_individuals} performs well for high \(N\).

@@

The payoff maximizing strategies typically will not defect before the opponent's
first defection, possibly because the training strategy collection contains some
strategies such as Grudger and Fool Me Once that retaliate harshly by defecting
for the remainder of the match if the opponent has more than a small number of
cumulative defections. Paradoxically it is advantageous to defect (as a signal)
in order to achieve mutual cooperation with opponents using the same strategy
but not with other opponents. Nevertheless an evolutionary process is able to
tunnel through the costs and risks associated with early defections to find more
optimal solutions, so it is not surprising in hindsight that handshaking
strategies emerge from the evolutionary training process.

@@

The strategies trained to maximize payoff in head-to-head matches are generally
cooperative and are effective invaders.
Combined with the fact that handshaking strategies are stronger resistors,
this suggests that while maximizing individual payoff can lead to the evolution
of cooperation, these strategies are not the most evolutionarily stable
in the long run. A strategy with a handshaking mechanism is still capable of
invading and is more resistant to subsequent invasions. Moreover, the
best resistor of the payoff maximally trained strategies (Evolved Looker Up
1\_1\_1),
which always defects if the opponent defects in the first round, is effectively
employing a one-shot handshake of C. Similarly, Grudger (also known as Grim),
which emerged from training memory one strategies for the Moran process,
also effectively employs a handshake of always cooperating, as it defects
for the remainder of the match if the opponent ever defects.

>>>>

This manuscript provides a detailed numerical analysis of agent-based simulations
of \textbf{\input{./tex/num_strategies.tex}}complex and adaptive strategies for the
IPD\@. This is made possible by the Axelrod library \cite{axelrodproject}, an
effort to provide software for reproducible research for the IPD\@. The library
now contains over \input{./tex/num_strategies_axelrod.tex}parameterized
strategies including classics like TitForTat and WinStayLoseShift, as well as
recent variants such as OmegaTFT, zero determinant and other memory one
strategies, strategies based on finite state machines, lookup tables, neural
networks, and other machine learning based strategies, and a collection of novel
strategies. Not all strategies have been considered for this study: excluded
are those that make use of knowledge of the number of turns in a match
and others that have a high
computational run time. The large number of strategies are available thanks to
the open source nature of the project with over 50 contributors from around the
world, made by programmers and researchers \cite{Knight2016}. Three of the considered
strategies are finite state machines trained specifically for Moran processes
(described further in Section~\ref{sec:strategies}).

@@

\begin{itemize}
	\item TF1 \(N=12\), 0\% noise.
	\item TF2 \(N=10\), 0\% noise.
	\item TF3 \(N=8\), 1\% noise.
\end{itemize}

Each matchup of players was run to fixation to estimate the absorption
probabilities. The trained algorithms
were run for fewer than 50 generations. Training data for this is available at
\cite{data}.

@@

Each strategy pair is run
with starting population distributions of $(1, N-1)$, $(N/2, N/2)$ and $(N-1 ,
1)$, for \(N\) from 2 through 14. The fixation probability is then empirically
computed for each combination of starting distribution and value of \(N\).  The
Axelrod library can carry out exact simulations of the Moran process. Since some
of the strategies have a high computational cost or are stochastic, samples are
taken from a large number of match outcomes for the pairs of players for use in
computing fitnesses in the Moran process. This approach was verified to agree
with unsampled calculations to a high degree of accuracy in specific cases.
This is described in Algorithms~\ref{alg:data_collection}
and~\ref{alg:moran_process}.

@@

It can be seen that apart from CS, none of the strategies
of Table~\ref{tbl:summary_top_2} perform well for \(N\in\{3, 7, 14\}\). The new
top performing strategies are:

@@

Table~\ref{tbl:top_resist} shows the top strategies when ranked
according to \(x_{N-1}\) for \(N\in\{3, 7, 14\}\).
Once again none of the short memory strategies from
Section~\ref{sec:two_individuals} perform well for high \(N\).

@@

The payoff maximizing strategies typically will not defect before the opponent's
first defection, possibly because the training strategy collection contains some
strategies such as Grudger and Fool Me Once that retaliate harshly by defecting
for the remainder of the match if the opponent has more than a small number of
cumulative defections. Paradoxically it is advantageous to defect (as a signal)
in order to achieve mutual cooperation with opponents using the same strategy
but not with other opponents. Nevertheless an evolutionary process is able to
tunnel through the costs and risks associated to early defections to find more
optimal solutions, so it is not surprising in hindsight that handshaking
strategies emerge from the evolutionary training process.

@@

The strategies trained to maximize payoff in head-to-head matches are generally
cooperative and are effective invaders.
Combined with the fact that handshaking strategies are stronger resisters,
this suggests that while maximizing individual payoff can lead to the evolution
of cooperation, these strategies are not the most evolutionarily stable
in the long run. A strategy with a handshaking mechanism is still capable of
invading and is more resistant to subsequent invasions. Moreover, the
best resistor of the payoff maximally trained strategies (Evolved Looker Up
1\_1\_1),
which always defects if the opponent defects in the first round, is effectively
employing a one-shot handshake of C. Similarly, Grudger (also known as Grim),
which emerged from training memory one strategies for the Moran process,
also effectively employs a handshake of always cooperating, as it defects
for the remainder of the match if the opponent ever defects.
